Sensitivity Analysis

### 2024-06-10 Article_Sensitivity_Analsysis_short-Chamomile.tex
10.06

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
This study aims to investigate the extraction process of caraway oil from chamomile flowers. A mathematical model is used to describe the governing mass transfer phenomena in a solid-fluid environment under supercritical conditions using carbon dioxide as a solvent. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. The physical properties of the solvent are estimated from the Peng-Robinson equation of state. A set of laboratory experiments was performed under multiple constant operating conditions: $30 - 40^\circ C$, $100 - 200$ bar, and $3.33-6.67 \cdot 10^{-5}$ kg/s. Different sensitivity analysis methods can be applied to assess the robustness of the model parameters and its influence on the process model. The local sensitivity analysis investigates the influence of infinitely small changes in the model parameters and the controls on model output. In particular,  this work focuses on analysing the step change of pressure on the extraction yield and the model state space.

@@

Sensitivity analysis is a tool to understand how parameters affect a model's output. In the case of dynamical systems, local sensitivity analysis provides a time series describing how that dependency evolves. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. The local sensitivity analysis techniques consider only a small region of parameter space, and the conclusions derived from such an analysis are limited to local conditions unless the discussed system is a linear model. The sensitivity equations can be obtained in various ways. This work implemented the automatic differentiation technique to derive the sensitivity equations. The effect of pressure increase on the outcome of the supercritical extraction model is analysed in this work. At given operating conditions ($35^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s), the step change of pressure enhance the mass transfer, which leads to faster loss of solute from solid particles and consequently to negative sensitivities in the solid phase.  Analogously, the sensitivities in the fluid phase are characterized by positive deviations, which indicates that more solute is transported to the fluid phase. As a result, the extraction yield is improved and characterised by positive sensitivities as well. Local sensitivity analysis can provide valuable information about process modelling, experiment design, or model reduction by identifying which parameters are influential and how these influence changes over time.

>>>>

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
This study investigates the process of caraway oil extraction from chamomile flowers. A parameter-distributed model consisting of a set of partial differential equations is used to describe the governing mass transfer phenomena in a solid-fluid environment under supercritical conditions using carbon dioxide as a solvent. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. The physical properties of the solvent are estimated from the Peng-Robinson equation of state. A set of laboratory experiments was performed under multiple constant operating conditions: $30 - 40^\circ C$, $100 - 200$ bar, and $3.33-6.67 \cdot 10^{-5}$ kg/s. Different sensitivity analysis methods can be applied to assess the robustness of the model parameters and their influence on the process model. The local sensitivity analysis investigates the influence of infinitely small changes in the model parameters and the controls on model output. This work focuses on analysing the effect of pressure on the model state space and the extraction yield.

@@

Sensitivity analysis is a tool to understand how model parameters affect its output. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. The local sensitivity analysis techniques consider only a small region of parameter space, and the conclusions derived from such an analysis are limited to local conditions. This work implemented the automatic differentiation to derive the sensitivity equations. In the case of dynamical systems, local sensitivity analysis provides a time series describing how that dependency evolves.

The local sensitivity analysis can be performed with respect to any model parameter, but this work focuses on the effect of pressure increase. At selected operating conditions ($35^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s), the pressure increase enhances the mass transfer, which leads to faster loss of solute from particles and consequently to negative sensitivities in the solid phase. Analogously, the sensitivities in the fluid phase are characterised by positive deviations, which indicates that more solute is transported to the fluid phase. As a result, the extraction yield is improved and characterised by positive sensitivities. Local sensitivity analysis can provide valuable information by identifying which parameters are influential and how these influence changes over time. The sensitivity analysis results can be utilised for process modelling, experiment design, or model reduction.

### 2024-06-10 Sections/Results_Sensitivity.tex
10.06

	This study examines the impact of pressure change on the state space and extraction yield. The process model and parameters are described in {\color{red}article 1}. Calibration of the process model was based on experiments performed under varying operating conditions: temperatures between $30 - 40^\circ C$, pressures between $100 - 200$ bar, and mass flow between $3.33-6.67 \cdot 10^{-5}$ kg/s. A sensitivity analysis was conducted, assuming the system operates at the midpoint of the validated range: $35^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s.

>>>>

	This study examines the impact of pressure change on the state space and extraction yield through the local sensitivity analysis. The details of the process model and parameters are delivered in {\color{red}article 1}. The process model was validated under the following range of operating conditions: temperatures between $30 - 40^\circ C$, pressures between $100 - 200$ bar, and mass flow rates between $3.33-6.67 \cdot 10^{-5}$ kg/s. The system is assumed to operate at the midpoint of the validated range: $35^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s.

### 2024-06-10 Sections/Sensitivity_Analysis.tex
10.06

	Following the work of \citet{Maly1996}, the sensitivity analysis equations(${\dot{Z}}$) can be defined as follow:

	{\footnotesize
		\begin{equation}
			\label{SA_def}
			Z(x;\Theta) = \cfrac{dx}{d\Theta}
	   \end{equation} }

	The new system of equations can be obtained by taking derivatives of $Z$ with respect to time $t$ and applying the chain rule.

	{\footnotesize
		\begin{equation} \label{SA_dt} 
			{\dot{Z}}(x;\Theta)  = \cfrac{dZ(x;\Theta)}{d t} = \cfrac{d}{d t} \left( \cfrac{dx}{d\Theta} \right) = \cfrac{d }{d\Theta} \left( \cfrac{dx}{d t} \right) = \cfrac{d {G}(x;\Theta)}{d\Theta} 
	\end{equation} }

@@

	{\footnotesize
		\begin{equation} \label{SA_eq_full}
			\cfrac{dG(x;\Theta)}{d\Theta} = \underbrace{ \cfrac{\partial {G}(x;\Theta)}{\partial x} }_{{J_x}(x;\Theta)} \underbrace{\cfrac{\partial x}{\partial \Theta} }_{S(x;\Theta)} + \underbrace{ \cfrac{\partial {G}(x;\Theta)}{\partial \Theta} }_{{J_\Theta}(x;\Theta)}
	  \end{equation} }

>>>>

	Following the work of \citet{Maly1996}, the sensitivity equations (${\dot{S}}$) can be defined as follow:

	{\footnotesize
		\begin{equation}
			\label{SA_def}
			S(x;\Theta) = \cfrac{dx}{d\Theta}
	   \end{equation} }

	The new system of equations can be obtained by taking derivatives of $S$ with respect to time $t$ and applying the chain rule.

	{\footnotesize
		\begin{equation} \label{SA_dt} 
			{\dot{S}}(x;\Theta)  = \cfrac{dS(x;\Theta)}{d t} = \cfrac{d}{d t} \left( \cfrac{dx}{d\Theta} \right) = \cfrac{d }{d\Theta} \left( \cfrac{dx}{d t} \right) = \cfrac{d {G}(x;\Theta)}{d\Theta} 
	\end{equation} }

@@

	{\footnotesize
		\begin{equation} \label{SA_eq_full}
			\cfrac{d G(x;\Theta)}{d \Theta} = \underbrace{ \cfrac{\partial {G}(x;\Theta)}{\partial x} }_{{J_x}(x;\Theta)} \underbrace{\cfrac{\partial x}{\partial \Theta} }_{S(x;\Theta)} + \underbrace{ \cfrac{\partial {G}(x;\Theta)}{\partial \Theta} }_{{J_\Theta}(x;\Theta)}
	  \end{equation} }

### 2024-06-10 Sections/Sensitivity_Pressure.tex
10.06

	A step-function shown in Figure \ref{fig:Sensitivty_P_P} illustrates the pressure change in the extractor. As discussed in Chapter \ref{CH:Governing_equations_chapter}, a small pressure wave propagates at the speed of sound relative to the flow. If the flow velocity is relatively low, all pressure changes are hydrodynamic (resulting from velocity motion) rather than thermodynamic. The Low Mach number assumption leads to instant propagation of the thermodynamic pressure throughout the system and allows for considering a single pressure value for the entire system. The model is built such that the pressure change first affects $\frac{dP}{dt}$; then in the next time step, the energy balance is influenced through $\frac{\partial (P A_f)}{\partial t}$. Analogously, the pressure change affects the enthalpy balance by introducing a step-change along the entire system as presented in Figure \ref{fig:Sensitivty_P_H}.

	\begin{figure}[h!]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/P_P_1.png}
		\caption{The effect of $P_{in}$ change on $P$ in the system}
		\label{fig:Sensitivty_P_P}
	\end{figure}

	The pressure change affects the fluid's temperature inside the computational domain, but boundary values are subject to constraints specified at the extremes of that domain. Applying the Dirichlet boundary conditions imposes a fixed temperature value at the inlet, which leads to a thermal gradient propagating along the system. In contrast, Neumann boundary conditions would dictate the heat flux at the boundaries. In this work, the Neumann boundary conditions equal to zero were applied to ensure that the temperatures at the inlet, the outlet, and the middle of the extractor vary, respectively, to the pressure change.

	\begin{figure}[h!]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/H_P_1.png}
		\caption{The effect of $P_{in}$ change on $(h \times \rho)$ in the system}
		\label{fig:Sensitivty_P_H}
	\end{figure}

	Figure \ref{fig:Sensitivty_P_CS} shows the sensitivity of the solute concentration in the solid phase in response to pressure change. As discussed in Chapter \ref{CH: Continuity}, the velocity of a fluid is inversely proportional to its density, which suggests that the pressure reduces the fluid's velocity. As a result, the residence time is extended, causing longer interaction between the solute and solvent. Initially, the extraction process is in the kinetic-controlled regime, where the concentration gradient is high, and the limiting factor is the solute solubility. As discussed in ({\color{red}article 1}), the system is considered to be far from saturation, which can explain the low system response at the beginning of the process. \citet{Fiori_2007} in their work also observed low sensitivity at the beginning of extraction. The system response becomes more evident when the concentration gradient starts diminishing, and the extraction switches from the kinetic-controlled to the diffusion-controlled regime. The negative sign can be interpreted as a faster solute loss from the solid phase, which corresponds to enhanced mass transfer. Over time, the amount of solute becomes a limiting factor, and the pressure change effect on the system is reduced. Eventually, the sensitivities approach zero asymptotically when the solute is exhausted from the solid bed.

@@

	Figure \ref{fig:Sensitivty_P_CF} shows how sensitive is the solute concentration in the fluid phase to the pressure change. Compared to Figure \ref{fig:Sensitivty_P_CS}, the dynamic behaviour of the fluid phase sensitivities can be observed. Due to the advection, the sensitivities move along the system analogously to the solute in the fluid phase.	Although the pressure increase enhances the mass transfer, the system response is initially low, which reflects the idle period discussed above. Later, the sensitivities increase due to faster solute loss from the solid phase. The positive sensitivities indicate that more solute is transported to the fluid phase. When the amount of solute in the solid phase becomes a limiting factor, the extraction rate slows down, and sensitivities decline asymptotically towards zero.

@@

	Figure \ref{fig:Sensitivty_P_y} illustrates how sensitive the extraction yield is to the pressure change. Initially, the sensitivity curve stays almost flat, suggesting a latency in the system's response to pressure changes. Due to the decreased velocity of the fluid, the solute reaches the extractor's outlet later, which causes minor negative sensitivities to appear. The process continues, and the sensitivity curve increases rapidly when the solute reaches the extractor's outlet. The positive yield sensitivity indicates improved process efficiency, which is directly related to enhanced mass transfer. The peak in $\frac{dy}{dP_{in}}$ reflects when the deviation from the original system is the largest. Eventually, the sensitivity declines and converges towards zero. The concentration gradient becomes a limiting factor, and the enhanced mass transfer no longer plays a dominant role compared to the original system.

>>>>

	A step-function shown in Figure \ref{fig:Sensitivty_P_P} illustrates the pressure change in the extractor. As discussed in Chapter \ref{CH:Governing_equations_chapter}, a small pressure wave propagates at the speed of sound relative to the flow. If the flow velocity is relatively low, all pressure changes are hydrodynamic (resulting from velocity motion) rather than thermodynamic. The Low Mach number assumption leads to instant propagation of the thermodynamic pressure throughout the system and allows for considering a single pressure value for the entire system. Analogously to Figure \ref{fig:Sensitivty_P_P}, the pressure change affects the enthalpy balance by introducing a step-change along the whole system as presented in Figure \ref{fig:Sensitivty_P_H}.

	\begin{figure}[h!]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/P_P_1.png}
		\caption{The effect of $P_{in}$ change on $P$}
		\label{fig:Sensitivty_P_P}
	\end{figure}

	The pressure change affects the fluid's temperature inside the computational domain, but boundary values are subject to constraints specified at the extremes of that domain. Applying the Dirichlet boundary conditions imposes a fixed temperature value at the inlet, which leads to a thermal gradient propagating along the system. In contrast, Neumann boundary conditions dictate the heat flux at the boundaries. In this work, Neumann boundary conditions of zero were used to ensure that the temperatures at the inlet, outlet, and inside the extractor would vary uniformly and according to the pressure change.

	\begin{figure}[h!]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/H_P_1.png}
		\caption{The effect of $P_{in}$ change on $(h \times \rho)$}
		\label{fig:Sensitivty_P_H}
	\end{figure}

	Figure \ref{fig:Sensitivty_P_CS} shows the sensitivity of the solute concentration in the solid phase in response to pressure change. As discussed in Chapter \ref{CH: Continuity}, the velocity of a fluid is inversely proportional to its density, which suggests that the pressure reduces the fluid's velocity. As a result, the residence time is extended, causing longer interaction between the solute and solvent. Initially, the extraction process is in the kinetic-controlled regime, where the concentration gradient is high, and the limiting factor is the solute solubility. As discussed in {\color{red}article 1}, the system is considered to be far from saturation, which explains the low system response at the beginning of the process. \citet{Fiori_2007} also observed low sensitivity at the beginning of extraction. The system response becomes evident when the concentration gradient diminishes, and the extraction switches from the kinetic-controlled to the diffusion-controlled regime. The negative sign can be interpreted as a faster solute loss from the solid phase, which corresponds to enhanced mass transfer. Over time, the amount of solute becomes a limiting factor, and the pressure change effect on the system is reduced. Eventually, the sensitivities approach zero asymptotically when the solute is washed out from the solid bed.

@@

	Figure \ref{fig:Sensitivty_P_CF} shows how sensitive is the solute concentration in the fluid phase to the pressure change. Compared to Figure \ref{fig:Sensitivty_P_CS}, the dynamic behaviour of the fluid phase sensitivities can be observed. Due to the advection, the sensitivities move along the system analogously to the solute in the fluid phase. Although the pressure increase improves the mass transfer, the system response is initially low, which reflects the idle period discussed above. Later, the sensitivities increase due to faster solute loss from the solid phase. The positive sensitivities indicate that more solute is transported from the particles to the fluid. When the amount of solute in the solid phase becomes a limiting factor, the extraction rate slows down, and sensitivities decline asymptotically towards zero.

@@

	Figure \ref{fig:Sensitivty_P_y} illustrates how sensitive the extraction yield is to the pressure change. Initially, the sensitivity curve stays almost flat, suggesting a latency in the system's response. Due to the decreased velocity of the fluid, the solute reaches the extractor's outlet later, which causes minor negative sensitivities to appear. When the solute reaches exits the extractor, the sensitivity curve increases rapidly. The positive yield sensitivity indicates improved process efficiency, which is directly related to enhanced mass transfer. The peak in $\frac{dy}{dP_{in}}$ reflects when the deviation from the original system is the largest. Eventually, the sensitivity declines and converges towards zero. The concentration gradient becomes a limiting factor, and the enhanced mass transfer no longer plays a dominant role compared to the original system.
